Reflections on reflection for \\ intuitionistic set theories

### 2019-01-12 paper-reflection.tex
typo

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

@@

\title[]{Reflections on reflection for \\ intuitionistic set theories}
\author{}
\email{}

@@

The reflection principle expresses that truth of any formula can already be
checked in an initial segment of the universe. This observation
is important both for philosophical and for practical
reasons: Philosophically, it tells us that the set-theoretic universe cannot be
distinguished from its initial segments by any set-theoretical property.
Practically, it allows to
transfer results obtained for a restricted class of objects to all such
objects. For instance, if we manage to verify (the $S$-relativization of) some
group-theoretic statement for all groups contained in an arbitrary set~$X$,
then we may deduce by the reflection principle that the statement holds
for all groups in the universe. Examples from sheaf cohomology and more
generally category theory abound; an example from set theory is presented in
Proposition~\ref{prop:zf-rrs2}.

@@

Aczel's $\RRS$ should not in itself be regarded as a choice principle; however in
its presence, ordinary~$\DC$ entails~$\RDC$. This result is due to
Aczel~\cite[Theorem~2.4]{aczel:rrs}, who proved the equivalence~$\RDC = \RRS +
\DC$ over~$\CZF^-$, and we give a new proof of this equivalence over the
stronger theory~$\IZF$, in Section~\ref{sect:appl-rdc}.

@@

\begin{thm}\label{thm:izf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory. Then~$\IZF+\RRS_2$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
Furthermore, the resulting set~$S$ may be supposed to be transitive and to be closed
under subsets; and given not a single formula~$\varphi$ but a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
all of them.
\end{thm}

@@

An alternative is offered by~$\ZFCI$, $\ZFC$ plus the existence of a
strongly inaccessible cardinal. In this system, there is a Grothendieck
universe~$U$; we can form the category~$\Set_U$ of all sets of~$U$ as an honest
set; can define truth in~$\Set_U$; and prove, within the system, that for any
formula~$\varphi$, $(\Set_U \models \varphi) \Leftrightarrow (U \models
\varphi)$. This even holds for formulas of the full infinitary language of
toposes, which allows infinite disjunctions and infinite conjunctions; this
extended language could not be treated by resorting to~$\PRA$ as indicated above.

However, truth in a Grothendieck universe~$U$ need not be related to actual
truth. A solution to this problem is provided by~$\ZFCS$ and by~$\ES$. In these systems, we can
form the topos~$\Set_\SS$, define truth in it, prove for all formulas~$\varphi$
that~$(\Set_\SS \models \varphi) \Leftrightarrow (\SS \models \varphi)$; and
reflection for~$\SS$ ensures that for each (external, standard)
formula~$\varphi$, the system proves~``$(\Set_\SS \models \varphi)
\Leftrightarrow \varphi$''.

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsthm,mathtools,stmaryrd,amssymb,graphicx}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

@@

\title[Reflections on reflection for intuitionistic set theories]{Reflections on reflection for \\ intuitionistic set theories}
\author{}
\email{}

@@

The reflection principle expresses that truth of any formula can already be
checked in an initial segment of the universe. This observation
is important both for philosophical and for practical
reasons: Philosophically, it tells us that the set-theoretic universe cannot be
distinguished from its initial segments by any set-theoretical property.
Practically, it allows to
transfer results obtained for a restricted class of objects to all such
objects. For instance, if we manage to verify (the $S$-relativization of) some
group-theoretic statement for all groups contained in an arbitrary set~$S$,
then we may deduce that the statement holds
for all groups in the universe. Examples from sheaf cohomology and more
generally category theory abound; an example from set theory is presented in
Proposition~\ref{prop:zf-rrs2}.

@@

Aczel's $\RRS$ should not in itself be regarded as a choice principle; however in
its presence, ordinary~$\DC$ entails~$\RDC$. This result is due to
Aczel~\cite[Theorem~2.4]{aczel:rrs}, who proved the equivalence~$\RDC = \RRS +
\DC$ over~$\CZF^-$, and we give a new proof of this equivalence over the
stronger theory~$\IZF$ in Section~\ref{sect:appl-rdc}.

@@

\begin{thm}\label{thm:izf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory. Then~$\IZF+\RRS_2$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n). \]
Furthermore, the resulting set~$S$ may be supposed to be transitive and to be closed
under subsets; and given not a single formula~$\varphi$ but a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
all of them.
\end{thm}

@@

An alternative is offered by~$\ZFCI$, $\ZFC$ plus the existence of a
strongly inaccessible cardinal. In this system, there is a Grothendieck
universe~$U$; we can form the category~$\Set_U$ of all sets in~$U$ as an honest
set; can define truth in~$\Set_U$; and prove, within the system, that for any
formula~$\varphi$, $(\Set_U \models \varphi) \Leftrightarrow (U \models
\varphi)$. This even holds for formulas of the full infinitary language of
toposes, which allows infinite disjunctions and infinite conjunctions; this
extended language could not be treated by resorting to~$\PRA$ as indicated above.

However, truth in a Grothendieck universe~$U$ need not be related to actual
truth. A solution to this problem is provided by~$\ZFCS$ and by~$\ES$. In these systems, we can
form the topos~$\Set_\SS$ of all sets in~$\SS$, define truth in it, prove for all formulas~$\varphi$
that~$(\Set_\SS \models \varphi) \Leftrightarrow (\SS \models \varphi)$; and
reflection for~$\SS$ ensures that for each (external, standard)
formula~$\varphi$, the system proves~``$(\Set_\SS \models \varphi)
\Leftrightarrow \varphi$''.
